---
title: "Does Left Handed Pitching Matter"
author: "Carl Ganz"
date: "September 26, 2015"
output: pdf_document
---

I want to explore whether more left handed pitching is better. Let's manipulate `Lahman's` baseball data before we do any modeling.

```{r warning=F}
#load packages
library(Lahman)
library(dplyr)
library(magrittr)
library(ggplot2)
#load datasets as datatables
pitching<-tbl_df(Pitching)
master<-tbl_df(Master)
teams<-tbl_df(Teams)
#select lefthandedness from master table
master<-select(master,playerID,throws)
#filter data for post war era
pitching<-filter(pitching,yearID>1945 & yearID<1980)
teams<-filter(teams,yearID>1945 & yearID<1980)
#merge in handedness with pitching data
pitching<-merge(pitching,master,by="playerID")
#generate summary statistics for teams: number of innings pitched by lefties 
#and number of games started by lefties for each team each year
leftyip<-pitching %>% group_by(yearID,teamID) %>% 
  filter(throws=="L") %>% summarise(leftyips=sum(IPouts),leftystarts=sum(GS))
#merge summary statistics in with team statistics
teams<-merge(teams,leftyip,by=c("yearID","teamID"))
#calculate percentage of innings pitched by lefties
teams$leftypercent<-100*teams$leftyips/teams$IPouts                              

```

Now lets use linear regressions to model how lefthandness correlates with winning.

```{r}
lm1<-lm(W~leftypercent,teams)
summary(lm1)

ggplot(teams,aes(x=leftypercent,y=W)) + geom_point(shape=2,alpha=1/2) + geom_smooth(method=lm)
```

We get a statistically significant positive correlation (each additional percent of innings thrown by left hander is associated with an extra $0.09$ wins per season), but as the graph and the $R^2$ both indicate the proportion of innings pitched by lefties isn't a good predictor of team performance.

Let's look at the number of games started by lefties instead.

```{r warning=F}
lm2<-lm(W~leftystarts,teams)
summary(lm2)

ggplot(teams,aes(x=leftystarts,y=W)) + geom_point(shape=2,alpha=1/2) + geom_smooth(method=lm)
```

We get similar results. Each additional start by a lefty is associated with an additional $0.06$ wins per season.

Let's instead model ERA instead of wins

```{r warning=F}
lm3<-lm(ERA~leftypercent,teams)
summary(lm3)

ggplot(teams,aes(x=leftypercent,y=ERA)) + geom_point(shape=2,alpha=1/2) + geom_smooth(method=lm)

lm4<-lm(ERA~leftystarts,teams)
summary(lm4)

ggplot(teams,aes(x=leftystarts,y=ERA)) + geom_point(shape=2,alpha=1/2) + geom_smooth(method=lm)
```

Each additional percent of innings pitched by a lefty is associated with $0.006$ decline in ERA over the season. Each additional start by a lefty is associated with $0.002$ decline in ERA.